An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-07-19 nips_2016.tex
more edits

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to perform
Metropolis-Hastings tests on the full data. Previous work on reducing
the cost of Metropolis-Hastings tests yield variable data reductions
and in the worst case require the entire dataset.  Here we present a
simple test which requires constant size minibatches if an auxiliary
condition is met. We describe several cases under which this condition
can be satisfied. Our new test is naturally tolerant of ``noisy''
estimates of the transition probability ratio. To achieve this, we
replace the classical Metropolis-Hastings acceptance test with a
smooth acceptance function which is a sigmoid (logistic)
function of the log likelihood ratio.
The resulting test can be combined with minibatch proposals to yield
updates with the same complexity as a simple SGD update. In this paper
we derive the test, discuss its implementation, and present several
experiments.
\end{abstract}

@@

Here we derive a new test which incorporates the variability in
minibatch statistics as {\em a natural part of the test}. Because of
this, the amount of data required for each test is fixed with high
probability and in expected value. Our test requires a bound on the
(relative) variance of the log acceptance probability. This variance
is small when the proposal distribution is efficient, that is when samples
diffuse at an optimal rate.

The variance of the log acceptance probability can be controlled in
three ways, and these allow the test to be applied to modest-sized
minibatches. The first two allow the test to be applied when the
minibatch size is fixed a-priori.

@@

\item Increase the temperature of the target distribution. Log likelihoods
  scale as $1/T$, and so the variance of the likelihood ratio will
  vary as $1/T^2$.

\item For continuous probability distributions, reduce the proposal
  step size and variance (for stochastic proposals).

\item Increase the minibatch size. Relative variance scales as $1/k$ vs the
  minibatch size $k$. Increased minibatch size also reduces the error of
  the test.

@@

@@

\begin{itemize}
\item We develop a new, more efficient (in samples per test) minibatch acceptance test which incorporates
online  estimates of its error probability.
\item We analyze the test for accuracy and speed.
\item We compared performance of our new test and prior approaches on several datasets. We demonstrate the test is several orders
  of magnitude more efficient than prior work measured as data consumed per test, and that it does not suffer
  from long-tailed data requirements (up to the dataset size). 
\end{itemize}

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$, we
can write
\begin{equation}\label{eq:relationship}
\Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is approximately
normal with variance $\sigma^2(\Delta^*)$. 
But to perform the test in equation (\ref{theexacttest}) we want $\Delta + X$ for
a logistic random $X$ (call it $X_{\rm log}$ from now on). 
Our approach is to decompose $X_{\rm log}$ as
\begin{equation}\label{eq:deconvolution}
X_{\rm log} = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ has zero-mean, Normal distribution with
variance $\sigma^2$, and $X_{\rm corr}$ is a zero-mean ``correction''
term with density $C_{\sigma}(X)$.  The two variables are added
(i.e., their distributions convolve) to form $X_{\rm log}$. It will be
convenient to express this relationship in terms of the CDFs $N_{\sigma}(X_{\rm
  norm}) = \Phi(X_{\rm norm}/\sigma)$ and $S(X_{\rm log})$, where
$\Phi(.)$ is the Normal CDF with zero mean and variance $\sigma^2$ and
$S(.)$ is the standard logistic function.

@@

>>>>

\begin{abstract}
Markov chain Monte Carlo (MCMC) methods have many applications in
machine learning. We are particularly interested in their application
to modeling very large datasets, where it is impractical to perform
Metropolis-Hastings tests on the full data. Previous work on reducing
the cost of Metropolis-Hastings tests yield variable data consumed per
sample, with only constant factor reductions vs. using the full
dataset for each sample.  Here we present a method that can be tuned
to provide arbitrarily small batch sizes, by adjusting either proposal
step size or temperature. Our approach uses the natural noise present
in minibatch likelihood esimates to furnish the randomness in a
Metropolis-Hastings test. Our test uses the noise-tolerant Barker
acceptance test with a novel additive correction variable.  The
resulting test can be combined with minibatch proposals to yield
updates with the same complexity as a simple SGD update. In this paper
we derive the test, analyze its performance, discuss its
implementation, and present several experiments.
\end{abstract}

@@

Here we derive a new test which incorporates the variability in
minibatch statistics as {\em a natural part of the test}. Because of
this, the amount of data required for each test is fixed with high
probability and in expected value. We use a Barker
test function \cite{Barker65} rather than a Metropolis test, which
makes the test naturally error tolerant. Similar ideas were presented
in \cite{Bardenet15} section 6.3. But the asymptotic test statistic
CDF and the Barker function are different, which leads to fixed
errors for the approach in \cite{Bardenet15}. Here we show that the
difference between the distributions can be corrected with an
additive random variable. This leads to a test which is fast, and
whose error can be made arbitrarily small.

Our test is applicable when the variance (over data samples) of the
log acceptance probability is small enough (less than 1). Its not
clear at first why this quantity should be so bounded. But we will see
that it is ``natural'' for well-specified models running
Metroplis-Hastings sampling with optimal proposals
\cite{OptimalScaling01} on a full dataset. When we reduce the amount
of data for the test, the variance goes up, and we have to reduce
it in one of several ways. Either:

@@

\item Increase the temperature of the target distribution. Log likelihoods
  scale as $1/T$, and so the variance of the likelihood ratio will
  vary as $1/T^2$. Our model is no longer well-specified (we are doing inference
  at a temperature different from that assumed during data generation), but
  higher temperature can be advantageous for parameter exploration.

\item For continuous probability distributions, reduce the proposal
  step size and variance (for stochastic proposals) compared to an optimal
  proposal. The variance of the log acceptance probability scales as the
  square of proposal step size.

\item Increase the minibatch size. log acceptance variance scales as $1/k$ vs the
  minibatch size $k$. Increased minibatch size also reduces the error rate for
  the test.

@@

Its worth discussing at this point what are the goals (typically) of
M-H sampling on very large datasets.  By the Bernstein-von Mises
theorem, the posterior distribution of the parameter $\theta$ for a
Bayesian inference task is asymptotically normal, and has variance
that scales inversely with the number of data samples $N$. This mode
is extremely sharp for large datasets, which may contain millions or
billions of samples. Simply sampling from this distribution is one
application, but an efficient proposal distribution
\cite{OptimalScaling01} has similar variance to the target
distribution and will diffuse to it extremely slowly from an random
initialization value. If there are any other strong modes, it is very
likely for the sampler to find one of them and become trapped in it
when run at the normal distribution temperature (T=1). A common
solution is to anneal the sampler, running first at high temperature
(scaling log likelihoods by 1/T) which flattens the likelihood
landscape.  This is turn reduces the variance of the log acceptance
probability and allows our acceptance test to be applied.

A second question concerns step size. Once we have fixed temperature,
our variance constraint implies that we have to trade-off proposal
step size $s$ and batch size $b$ ($b \propto p^2$). i.e. we can make
many small steps, or one large step, with a given batch of data. One
of the primary drivers of this work is our belief in the value of
small steps. For applications to neural networks or other models where
the posterior is multimodel, posterior inference is arguably a search
process. Covering the search space densely with small steps is much
more valuable than few sparse steps toward the nearest optimum.

@@

\begin{itemize}
\item We develop a new, more efficient (in samples per test) minibatch acceptance test with quantifiable error bounds. The test
  uses a novel additive correction variable to implement a Barker test based on minibatch mean and variance. 
\item We analyze the test for accuracy and speed.
\item We compared performance of our new test and prior approaches on several datasets. We demonstrate the test is several orders
  of magnitude more efficient than prior work measured as data consumed per test, and that it does not suffer
  from long-tailed batch sizes (up to the dataset size). 
\end{itemize}

@@

In the limit, since $\Delta^*$ is normally distributed about its mean $\Delta$, we
can write
\begin{equation}\label{eq:relationship}
\Delta^* = \Delta + X_{\rm norm}, \quad X_{\rm norm} \sim \mathcal{\bar{N}}(0, \sigma^2(\Delta^*)),
\end{equation}
where $\mathcal{\bar{N}}(0, \sigma^2(\Delta^*))$ denotes a distribution which is approximately
normal with variance $\sigma^2(\Delta^*)$. 
But to perform the test in equation (\ref{theexacttest}) we want $\Delta + X$ for
a logistic random $X$ (call it $X_{\rm log}$ from now on). 
Our approach is to decompose $X_{\rm log}$ as
\begin{equation}\label{eq:deconvolution}
X_{\rm log} = X_{\rm norm}+X_{\rm corr},
\end{equation}
where we assume $X_{\rm norm}$ has zero-mean, Normal distribution with
variance $\sigma^2$, and $X_{\rm corr}$ is a zero-mean ``correction''
term with density $C_{\sigma}(X)$.  The two variables are added
(i.e., their distributions convolve) to form $X_{\rm log}$. It will be
convenient to express this relationship in terms of the CDFs $N_{\sigma}(X_{\rm
  norm}) = \Phi(X_{\rm norm}/\sigma)$ and $S(X_{\rm log})$, where
$\Phi(.)$ is the Normal CDF with zero mean and unit variance and
$S(.)$ is the standard logistic function.

@@

\section{Computing the Correction Distribution}\label{sec:correction}
Our goal is to compute the distribution of the correction variable $X_{\rm corr}$ such that
$X_{\rm norm} + X_{\rm corr} = X_{\rm log}$, where $X_{\rm norm} \in \mathcal{N}(0,\sigma)$
and $X_{\rm log}$ has a standard logistic CDF $(1+\exp(-X))^{-1}$.
In the next section we show that the accuracy of the test depends on the absolute
error between the CDFs of $X_{\rm norm} + X_{\rm corr}$ and the distribution of $X_{\rm log}$.
So we minimize this in our construction.

Our goal then is to minimize
\begin{equation}
  |\Phi_{\sigma} * C_{\sigma} - S|
\end{equation}
where $\Phi_{\sigma}(X) = \Phi(X/\sigma)$ and $\Phi$ is the standard
normal CDF, $S(X)$ is the logistic function, $C_{\sigma}(X)$ is the
{\em density} of the correction distribution, and $*$ denotes
convolution. In continuous form, this value is
\begin{equation}
  \label{eq:conv1}
  \sup_X\left|\int_{-V}^{V}\Phi_{\sigma}(X-Y) C_{\sigma}(Y) dY - S(X)\right|
  \end{equation}
Assuming $X_{\rm corr}$ lies in the interval $[-V,V]$. For computation of $C_{\sigma}(Y)$, we
discretize $X$ and $Y$ into $4N+1$ and $2N+1$ values respectively, with $X \in [-2V,2V]$ and
$Y \in [-V,V]$. Writing $X_i = i * V / N$ and $Y_j = j * V / N$, the condition in
equation (\ref{eq:conv1}) can be written:
\begin{equation}
  \max_{i \in \{-2N,\ldots,2N\}}\left|\sum_{j = -N}^{N}\Phi_{\sigma}(X_i-Y_j) C_{\sigma}(Y_j) - S(X_i)\right|
\end{equation}
and defining a matrix $M_{ij} = \Phi_{\sigma}(X_i-Y_j)$, vectors $U_j = C_{\sigma}(Y_j)$ and $V_i = S(X_i)$, the
above is equivalent to minimizing the 1-norm:
\begin{equation}
  \label{eq:error_norm}
  ||MU-V||_1
\end{equation}
wrt $U$. We have an additional constraint that $U_j > 0$ for all $j$, since $U$ represents
a density. We first explored optimizing this system with linear programming to find the $U$ such that:
\begin{equation}
  \begin{split}
    & \min(\epsilon) \qquad \hbox{where} \\
    & -\epsilon \leq MU - V \leq \epsilon \\
    & U \geq 0
  \end{split}
\end{equation}
which was tractable up to a few hundred dimensions. However, the discretization error is bounded by
the curvature of the underlying functions which are here slightly less than one. i.e.
the errors are of the order of $(V/N)^2$. Here we chose $V=20$ to provide adequate containment of
the distributions (the CDFs are extremely close to either zero or one outside this range). So
with 200 points, we have a discretization error of approximately 0.01. This is quite high. To yield
higher resolution and lower error, we switched to a least squares solution. 
Thus we seek to find $U$ that minimizes
\begin{equation}
  ||MU-V||_2
\end{equation}
Its not clear that this will yield a good 1-norm for equation (\ref{eq:error_norm}), but in practice it does.
Because the solution of this system is not very stable, especially in high dimensions, we actually
minimize a regularized version
\begin{equation}
  ||MU-V||_2 + \lambda||U||_2
\end{equation}
the solution is
\begin{equation}
  U = (M^T M + \lambda I)^{-1}V
\end{equation}
with this approach, there is no guarantee that $U \geq 0$. However, we have some
flexibility in the choice of $\sigma$ in equation (\ref{eq:conv1}). As we decrease
the variance of $X_{\rm norm}$, the variance of $X_{\rm corr}$ grows by the same amount
and is in fact the result of convolution with a gaussian whose variance is the difference.
Thus as $\sigma$ decreases, $C(X)$ grows and approaches the derivative of a
logistic function at $\sigma = 0$. It retains some very weak negative values for $\sigma > 0$
but removal of those values leads to very small error.
\begin{table}[h]
  \centering
  \begin{tabular}{|l|l|l|l|}
    \hline
     N & $\sigma$ & $\lambda$ & L1 error  \\
        \hline
     4000 & 0.9 & 1 & 1.0e-4  \\
     4000 & 0.8 & 0.03 & 5.0e-6 \\
     \hline
    \end{tabular}
\end{table}
